M\"obius Inversion: From Posets to Categories

### 2017-04-12 Drafts/Draft.tex
with approximation

\newpage
\pagenumbering{arabic}
\begin{chapter}{Introduction to M\"obius Inversion}\label{chap_intro}
\section{Introduction and Motivation}
%TODO: Expand introduction with more explicit examples
We motivate M\"obius inversion using several examples.
\begin{examp}
Suppose that we have the integer sequence $g$ of ones:
\[
g : 1, 1, 1, \cdots
\]
We can let $f$ be the sequence of cumulative sums of $g$. So $f_k = g_1 + \cdots + g_k$. We can write down the values:
\[
f: 1,2,3,\cdots
\]
In general, summing up values is easy. Finding values of $f$ from $g$ was not difficult.

@@

The goal of M\"obius inversion is to generalize the process of reversing sums over discrete, ordered structures called \emph{posets}. If sums like the above are a discrete analogue of integration (cumulatively summing over a region), then M\"obius inversion is discrete differentiation. In the same way that developing the theory of the derivative is useful for solving differential equations where functions are defined in terms of their integral, M\"obius inversion helps us extract values which are defined in terms of their cumulative sums. In addition to building the necessary machinery for M\"obius inversion, we will consider several applications of it, including a somewhat surprising one in data clustering.
\section{Initial Example and Partially Ordered Sets}
M\"obius inversion is a general construct. The classical definition of a M\"obius function is a function $\mu$ on the positive integers which satisfies the following for functions $f,g$ on the positive integers:
\begin{equation}\label{init_mob} 
f(n) = \sum_{d | n} g(d) \iff g(n) = \sum_{d | n}  f(d)\mu\left(\frac dn\right)
\end{equation}

Our first goal is to prove the existence of, and find, this function $\mu$. We will do this using by developing M\"obius inversion in general and then applying it. We start with several definitions.

@@

Note that, because $X$ is lower finite, this sum has a finite number of terms, so is always finite-valued. Using the ``indicator'' function $\zeta$ from above, we can rewrite the definition of $f$ to remove the need for the bound on the sum:
\begin{equation}\label{sum_form_intro}
f(a) = \sum_{b \in X}g(b)\zeta(b,a)
\end{equation}
Each value of $f$, therefore, takes on precisely the structure of a dot product. We will use this structure to leverage linear algebra: 
\begin{itemize}
\item Totally order $X$ such that the partial order is respected;
\item Write $g$ as a row vector which contains the values $g$ takes on elements of $X$, with respect to the total ordering;
\item Write $\zeta$ as a matrix whose $i,j$ entry is equal to $\zeta(a_i, a_j)$, where $a_i, a_j$ are the $i$'th and $j$'th elements of $X$ under the total ordering.
\end{itemize}

@@

This definition of $\mu$ depends only on $\frac nd$, which means that we can abuse notation to write $\mu(\frac nd)$ instead of $\mu(d,n)$. This allows us to use Theorem \ref{mob_inv} to arrive at \eqref{init_mob}:
\[
f(n) = \sum_{d | n} g(d) \iff g(n) = \sum_{d | n}  f(d)\mu\left(\frac nd\right).
\]
\section{Euler's Totient}
Euler's Totient is a famous function in number theory, which counts the how many numbers are relatively prime to a particular value. We denote the function by $\varphi : \Zz^+ \rightarrow \Zz^+$, and its definition is:
\[
\varphi(n) = \#\{1 \leq m \leq n : (m,n) = 1\},
\]
where $(m,n)$ denotes the greatest common divisor of $m,n$.

@@

We start with an important fact:
\begin{thm}[Gauss]\label{gauss}
Let $n \in \Zz^+$ be given. Then 
\[
n = \sum_{d|n}\varphi(d).
\]
\end{thm}
There are a variety of ways of proving this fact. Here we include a concise proof using group theory.
\begin{proof}
For any integer $d$, $\varphi(d)$ counts the generators of $C_d$. Each element of $C_n$ generates exactly one cyclic subgroup, so counting the total number of generators of cyclic subgroups of $C_n$ is equivalent to counting the elements of $C_n$. Also, $C_n$ has exactly one cyclic subgroup for each divisor of $n$ (and no others). So the sum indeed counts the number of elements of $C_n$, of which there are $n$.
\end{proof}
This sum gives us a value we have access to ($n$) in terms of a sum of values we want (values of $\varphi$). This is a perfect situation to apply M\"obius inversion, and undo that sum. We apply \eqref{init_mob}, which tells us
\[
\varphi(n) = \sum_{d|n} d\mu\left(\frac nd\right).
\]
Recall that Theorem \ref{mu_div} tells us how to write down values of $\mu$. Depending on what information we have access to, this might be much easier than just counting the number of values relatively prime to $n$. This is a straightforward example of how M\"obius inversion's ability to reverse a sum can be useful.
\end{chapter}

\begin{chapter}{Groups and Orbits} \label{chap_groups}
\section{Background}
For this chapter, it is necessary to be familiar with a simple structure called a \emph{group}. In this section, we will define it and briefly establish the machinery we need without going into too much detail, omitting proofs. More detail on elementary group theory can be found in an abstract algebra textbook. Knowledge of groups is not needed for any other chapters in this investigation.

@@

@@

This implies that
\[
\mu = \zeta^{-1} = \begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}.
\]
Next, we find $\sigma$. Recall that $\sigma$ is the vector of subgroup sizes. So 
\[
\sigma = [4, 2, 1].
\]
Finally, note that $|C_4|=4$. So
\begin{align*}
\lambda &= \frac 14 \mu \sigma^T\\
&= \frac 14 
\begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}
\begin{bmatrix} 4 \\ 2 \\ 1 \end{bmatrix}\\
&= \frac 14 \begin{bmatrix} 2 \\ 1 \\ 1\end{bmatrix}
\end{align*}
This $\lambda$ is independent of any set $G$ might act on.

So let's try a particular action. Suppose we have a $2 \times 2$ chessboard, and we want to place some number of rooks on the board such that none of them attack one another. If one configuration can be obtained from the other by rotating the board (ignoring light and dark squares), then we wish to count them as the same configuration. The goal is to count how many distinct configurations there are, which cannot be obtained from each other by rotation.

We can first solve this problem by inspection, or by "brute force," by just enumerating the possibilities:
\begin{description}
\item[Zero rooks:] There is one option with zero rooks: The empty board.
\[
\includegraphics{../Figures/zero_rooks.pdf}
\]
\item[One rook:] We can put one rook on any of the four squares. However, we can rotate the board to obtain any of these from the others. So there is just one distinct configuration. \\
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/one_rook.pdf}}
\]
\item[Two rooks:] We can put rooks on opposite corners. There are two such pairs, but we can rotate the board and obtain either pair from the other. So, again, there is only one distinct possibility.
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/two_rooks.pdf}}
\]
\end{description}
So there are a total of three distinct arrangements of rooks. Now we can try to solve this problem using $\lambda$.

@@

Luckily, we know how to count orbits. All we have to do is find $f$. We go through the subgroups of $C_4$.
\begin{itemize}
\item $C_4$ only fixes the empty board. Any other arrangement can be changed by a rotation. So $f(1) = 1$.
\item The subgroup $\langle \alpha^2 \rangle$ contains the identity element as well as the $180^\circ$ rotation. This subgroup fixes both arrangements with two rooks, as well as the empty board. So $f(2) = 3$.
\item The subgroup $\langle e \rangle$ fixes all the arrangements. There are 2 ways to place two rooks, 4 ways to place one rook, and 1 way to place zero rooks. So $f(2) = 7$.
\end{itemize}
Then all we have to do is compute $\lambda f^T$.
\begin{align*}
\lambda f^T &= \frac 14 [2, 1, 1] \begin{bmatrix} 1 \\ 3 \\ 7\end{bmatrix} \\
&= \frac 14 12\\
&=3
\end{align*}
So we indeed find the correct answer. 
\end{examp}

In this example, it's not clear that using $\lambda$ actually saved work - in fact, it seemed to take longer. But if we were to increase $n$, we could still determine the number of distinct arrangements relatively easily, because $\lambda$ wouldn't change. Furthermore, in many situations (including this one) it is much easier to determine how many \emph{total} arrangements are fixed by a subgroup that it is to find the number of \emph{distinct} arrangements. As an example of that, we can look at another example, which is more difficult to solve by hand. 
\begin{examp}
Suppose we have a $4 \times 4$ board, and we would like to color each square one of $k$ colors. We would like to consider colorings identical if one can be obtained from the other by rotating the board. The goal is to count the number of distinct colorings.

@@

\begin{chapter}{Inclusion-Exclusion}
\section{Introduction}
We start by introducing the desired end result, the \emph{inclusion-exclusion principle}:
\begin{thm}[Inclusion-Exclusion]\label{iep}
Let $S$ be a finite set. Let $A_i \subseteq S$ for each $i \in I$ where $I$ is finite. Let $K \subseteq I$ be given. Then
\[
\left|\bigcup_{i \in K}A_i\right| = \sum_{\emptyset \neq J \subseteq K}(-1)^{|J| + 1}\left|\bigcap_{i \in J}A_j\right|
.\]
\end{thm}
This is a well-known fact. It allows us to find the size of unions of overlapping sets if we have information about intersections of those sets. There are many ways of proving the Inclusion-Exclusion Principle. %TODO: Explain why proving it with Mobius inversion is worthwhile

@@

The goal of the inclusion-exclusion principle is, given a subset $K$ of $I$, to find the size of the union of the sets $A_i$ for $i \in K$. The difficulty arises from counting elements multiple times. For example, if we knew that the $A_i$ were disjoint, we could just consider
\[
\sum_{i \in K}\left|A_i\right|.
\]
But, because intersections of the $A_i$ might be nonempty, this sum might count some elements multiple times. We must approach the problem more carefully. First, we define the following indicator function $\chi: s \rightarrow \ps(I)$ by:
\[
\chi(x) = \{i \in I : x \in A_i\}.
\]
So $\chi$ gives us the indices of the subsets of $S$ which contain $x$.  
Next, we define some functions defined on our poset, $\ps(I)$.
Let $J \subseteq I$ be given. Define $f,g,\hat{f}, \hat{g}:\ps(I) \rightarrow \Nn$ by

@@

Also, we define the \emph{complement} function/matrix $c$ as follows: %TODO: Change variable name?
\[
c(J,K) = \begin{cases} 1 & J = K^c \\ 0 & \text{otherwise}. \end{cases}
\]
Note that by definition, $c(J,K) = c(K,J)$. So $c$ is symmetric. We also observe that $c^2 = I$, so $c$ is self-inverse.

@@

It makes sense that we would use M\"obius inversion to prove the Inclusion-Exclusion principle. The difficult aspect of counting how many elements are in the union of the sets is avoiding double-counting. %TODO: Finish this

@@

\begin{chapter}{Euler Characteristic of a Category}\label{chap_euler}
\section{Introduction to Categories} \label{cat_defs}

@@

We present several examples.
\begin{examp}
Let $G$ be a group. Then we can represent $G$ as a category $C$ as follows: 
\begin{itemize}
\item Let $C_0 = \{*\}$, a single object
\item Let $C(*, *) = G$. In other words, let the set of morphisms from $*$ to $*$ be equal to the set of group elements. 
\item Use the group operation for multiplication. In other words, if $g,h \in C(*,*)$, then, because $g,h \in G$, we can write $gh$. So let $g \circ h = gh$. 
\end{itemize}
It is straightforward to check that $C$ is indeed a category. 
\end{examp}

@@

\begin{defn}\label{def_weighting}
A \emph{weighting} on a category $\Aa$ is a function $\omega$ on $\obj(\Aa)$ such that for all $a \in \Aa$,
\[
\sum_{b \in C}|C(a,b)|\omega(b) = 1,
\]
or, equivalently, if
\[
\sum_{b \in C}\zeta(a,b)\omega(b) = 1.
\]
\end{defn}

One of the first questions we ask when we see this definition is how we can write find valid weightings. Luckily, this definition looks familiar. We have a value we know (1) expressed as the sum of function we are seeking ($\omega$). So we can use M\"obius inversion! The situation is slightly more complicated, because we don't know for sure that $\zeta$ is invertible. We formalize that:

\begin{lemma}
\label{mobius_is_weighting}
If $C$ has M\"obius inversion, then the weighting $\omega_\mu$ defined by $\omega_\mu(a) = \sum_b \mu(a,b)$ is a weighting on $C$.
\end{lemma}
\begin{proof} For all $a \in C$,
\begin{align*}
\sum_b \zeta(a,b)\omega_\mu(b) &= \sum_b\left( \zeta(a,b)\sum_c \mu(b,c)\right)\\
&=\sum_b\sum_c\zeta(a,b)\mu(b,c)\\
&=\sum_c\left(\sum_b\zeta(a,b)\mu(b,c)\right)\\
&=\sum_c \delta(a,c)\\
&=1.
\end{align*}
\end{proof}

Note that even if $C$ does not admit M\"obius inversion, it might admit a weighting. For example, the category in Example \ref{2_pt} admits the weighting $\omega(a) = \omega(b) = 0.5$, even though it does not admit M\"obius inversion.

@@

\begin{defn}[Leinster, \cite{Lein1}]\label{cat_euler}
Let $\omega$ be a weighting on a finite category $A$. Then the \emph{Euler--Leinster characteristic}, or just the \emph{Euler characteristic}, of $A$, denoted by $\chi(A)$, is defined by
\[
\chi(A) = \sum_{a \in A} \omega(a)
\]
\end{defn}
Note that even if a category admits several weightings, they all lead to the same Euler characteristic by a simple sum rearrangement argument. Together with Lemma \ref{mobius_is_weighting}, this implies the following:
\begin{cor}\label{euler_mu}
If a finite category $A$ has a M\"obius inversion $\mu$, then
\[\chi(A) = \sum_{a \subseteq A} \sum_{b \subseteq A} \mu(a, b).
\]
\end{cor}
%TODO: Think of category w/ no euler char
\section{Simplicial Complexes}
Calling a new construction an Euler characteristic needs some intuitive justification. We will introduce a structure called a \textit{simplicial complex} to show that calling this value an Euler characteristic is reasonable.

@@

The notation for $\cat(\sigma)$ is a little tricky. If $\sigma$ is a simplicial complex, and $\alpha \in \sigma$ is a simplex, then we write $\cat(\alpha)$ to denote the object in $\cat(\sigma)$ corresponding with the simplex $\alpha \in \sigma$. We consider an example.
\begin{examp}
Let $\sigma = \{a, b, ab\}$. Then $\cat(\sigma)$ has three objects and five morphisms, including the three identity morphisms.
\begin{figure}[h]
\[
\includegraphics{../Figures/simp_cat.pdf}
\]
\caption{$\sigma$ and $\cat(\sigma)$. Recall that we omit identity morphisms when drawing a category.}
\end{figure}
\end{examp}

@@

There are many methods of hierarchal clustering. In the above example, explicit clusters are formed by grouping points which are near each other. Methods like this are common; see \cite{Carl1}. We will explore an alternate method of clustering, in which we do not explicitly group points, but instead estimates the number of clusters using a real, rather than integer, value. We will use a generalization of the Euler--Leinster characteristic from Chapter \ref{chap_euler}. We start by building some important definitions.

@@

We turn to several examples of monoidal categories.
\begin{examp}
The category $\catname{set}$ is the prototypical example of a monoidal category. In that case, the monoidal product $\otimes$ is the Cartesian product on sets (hence the notation $\otimes$ for the monoidal product), and $u = \{*\}$, some set with a single element.

>>>>

\newpage
\pagenumbering{arabic}
\begin{chapter}{Introduction to M\"obius Inversion}\label{chap_intro}
\section{Introduction and Motivation}
We motivate M\"obius inversion using several examples.
\begin{examp}
Suppose that we have the integer sequence $g$ of ones:
\[
g : 1, 1, 1, \cdots
\]
We can let $f$ be the sequence of cumulative sums of $g$. So $f_k = g_1 + \cdots + g_k$. We can write down the values:
\[
f: 1,2,3,\cdots
\]
In general, summing up values is easy. Finding values of $f$ from $g$ was not difficult.

@@

The goal of M\"obius inversion is to generalize the process of reversing sums over discrete, ordered structures called \emph{posets}. If sums like the above are a discrete analogue of integration (cumulatively summing over a region), then M\"obius inversion is discrete differentiation. In the same way that developing the theory of the derivative is useful for solving differential equations where functions are defined in terms of their integral, M\"obius inversion helps us extract values which are defined in terms of their cumulative sums. In addition to building the necessary machinery for M\"obius inversion, we will consider several applications of it, including a somewhat surprising one in data clustering.
\section{Initial Example and Partially Ordered Sets}
M\"obius inversion is a general construction. It refers to the process of using what is called a \emph{M\"obius function} to access individual terms of a sum over some structure. We would like the M\"obius function to depend only on that structure, and not on the specific values being added up.

The classical definition of a M\"obius function is a real-valued function $\mu$ on the positive integers which satisfies the following for any functions $f,g$:
\begin{equation}\label{init_mob} 
f(n) = \sum_{d | n} g(d) \iff g(n) = \sum_{d | n}  f(d)\mu\left(\frac dn\right).
\end{equation}

Then M\"obius inversion is simply the process of using $\mu$ to obtain values of $g$ from values of $f$. Without knowing anything about either function other than how they relate to each other, we are able to write down $g$ in terms of $f$.

In fact, this function $\mu$ does exist. We will prove that, and find $\mu$, by developing M\"obius inversion in general and then applying it. We start with several definitions.

@@

Note that, because $X$ is lower finite, this sum has a finite number of terms, so is always finite-valued. Using the ``indicator'' function $\zeta$ from above, we can rewrite the definition of $f$ to remove the need for the bound on the sum:
\begin{equation}\label{sum_form_intro}
f(a) = \sum_{b \in X}g(b)\zeta(b,a)
\end{equation}
Each value of $f$, therefore, takes on precisely the structure of a dot product. We will use this structure to leverage linear algebra: 
\begin{itemize}
\item Totally order $X$ such that the partial order is respected;
\item Write $g$ as a (possibly infinite) row vector which contains the values $g$ takes on elements of $X$, with respect to the total ordering;
\item Write $\zeta$ as a (possibly infinite) matrix whose $i,j$ entry is equal to $\zeta(a_i, a_j)$, where $a_i, a_j$ are the $i$'th and $j$'th elements of $X$ under the total ordering.
\end{itemize}

The matrices and vectors we construct in this way might be infinite, so we have to be careful. In this chapter, all possibly infinite matrices we consider will be upper triangular (it is clear from the definition of $\zeta$ that this is the case here). Each term of the product of two upper triangular matrices is can be calculated from a finite number of matrix elements, even if the factor matrices are infinite. The same is true when an upper triangular matrix is left-multiplied by a row vector. Finally, because our matrices will be upper triangular and have ones on the diagonal, they will automatically be invertible (see Lemma \ref{inv_ex}), and can be inverted even if infinite. In subsequent chapters, we will restrict our attention to M\"obius inversion which uses only finite matrices.

@@

This definition of $\mu$ depends only on $\frac nd$, which means that we can abuse notation to write $\mu(\frac nd)$ instead of $\mu(d,n)$. This allows us to use Theorem \ref{mob_inv} to arrive at \eqref{init_mob}:
\[
f(n) = \sum_{d | n} g(d) \iff g(n) = \sum_{d | n}  f(d)\mu\left(\frac nd\right).
\]
This is a powerful, versatile result in number theory. As an example, we will walk through how this application of M\"obius inversion is related to Euler's Totient function.
\begin{examp}[Euler's Totient]
One of the most fundamental functions of number theory is Euler's Totient, which counts the how many numbers are relatively prime to a particular value. We denote the function by $\varphi : \Zz^+ \rightarrow \Zz^+$, and its definition is:
\[
\varphi(n) = \#\{1 \leq m \leq n : (m,n) = 1\},
\]
where $(m,n)$ denotes the greatest common divisor of $m,n$.

@@

We start with an important fact:
\begin{thm}[Gauss]\label{gauss}
Let $n \in \Zz^+$ be given. Then 
\[
n = \sum_{d|n}\varphi(d).
\]
\end{thm}
There are a variety of ways of proving this fact. Here we include a concise proof using group theory.
\begin{proof}
For any integer $d$, $\varphi(d)$ counts the generators of $C_d$. Each element of $C_n$ generates exactly one cyclic subgroup, so counting the total number of generators of cyclic subgroups of $C_n$ is equivalent to counting the elements of $C_n$. Also, $C_n$ has exactly one cyclic subgroup for each divisor of $n$ (and no others). So the sum indeed counts the number of elements of $C_n$, of which there are $n$.
\end{proof}
This sum gives us a value we have access to ($n$) in terms of a sum of values we want (values of $\varphi$). This is a perfect situation to apply M\"obius inversion, and undo that sum. We apply \eqref{init_mob}, which tells us
\[
\varphi(n) = \sum_{d|n} d\mu\left(\frac nd\right).
\]
Recall that Theorem \ref{mu_div} tells us how to write down values of $\mu$. Depending on what information we have access to, this might be much easier than just counting the number of values relatively prime to $n$. This is a straightforward example of how M\"obius inversion's ability to reverse a sum can be useful.
\end{examp}
\end{chapter}

\begin{chapter}{Groups and Orbits} \label{chap:groups}
\section{Motivation and Definitions}
A \emph{group} is one of the simplest possible algebraic structures. It is a set with a binary operation, analogous to addition or multiplication. Groups are useful for encoding symmetries and permutations, using a construction called a \emph{group action}, which we will define shortly. After we build some machinery, we can consider a motivating problem, which this chapter will give us the tools to solve.

For this chapter, it is necessary to be familiar with groups, including some basic facts about them. In this section, we will define it and briefly establish the machinery we need without going into too much detail, omitting proofs. More detail on elementary group theory can be found in an abstract algebra textbook. Knowledge of groups is not needed for any other subsequent chapters.

@@

This is enough terminology for a motivating problem.

\begin{examp}\label{ex:coloring_init}
Suppose we have a $4 \times 4$ board, and we would like to color each of the 16 squares using one of $k$ possible colors. We want to count the possible colorings. However, the board is symmetric, and we would like to consider colorings equivalent if the board can be rotated to obtain one coloring from the other. So our goal is to count how many \emph{distinct} colorings there are, where two colorings are distinct if they never coincide even if the board is rotated.

Our first guess is $k^{16}$, the number of ways of coloring the board without regard for rotations. But we will end up counting some colorings multiple times. Dividing by 4 is no good, because some colorings are not counted four times but two times or once.

One solution is to let a particular group act on the set of $k^{16}$ initial colorings, and count the number of orbits. If the group properly captures the symmetry of the board, colorings will be in the same group if and only if they are equivalent. So we need to develop the technology to count the orbits of a group action. This will occupy the next several sections, and we will use the tools we develop to solve this particular problem in Example \ref{ex:coloring}.
\end{examp}

@@

This implies that
\[
\mu = \zeta^{-1} = \begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}.
\]
Next, we find $\sigma$. Recall that $\sigma$ is the vector of subgroup sizes. So 
\[
\sigma = [4, 2, 1].
\]
Finally, note that $|C_4|=4$. So
\begin{align*}
\lambda &= \frac 14 \mu \sigma^T\\
&= \frac 14 
\begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}
\begin{bmatrix} 4 \\ 2 \\ 1 \end{bmatrix}\\
&= \frac 14 \begin{bmatrix} 2 \\ 1 \\ 1\end{bmatrix}
\end{align*}
We determined $\lambda$ before even deciding what set $G$ might act on. So all the work we have done so far can be reused for a variety of actions.

Let's try a particular action. Suppose we have a $2 \times 2$ chessboard, and we want to place some number of rooks on the board such that none of them attack one another. If one configuration can be obtained from the other by rotating the board (ignoring light and dark squares), then we wish to count them as the same configuration. The goal is to count how many distinct configurations there are, which cannot be obtained from each other by rotation.

We can first solve this problem by inspection, or by "brute force," by just enumerating the possibilities. This way, when we solve it with $\lambda$, we can determine whether we found the right answer.
\begin{description}
\item[Zero rooks:] There is one option with zero rooks: The empty board.
\[
\includegraphics{../Figures/zero_rooks.pdf}
\]
\item[One rook:] We can put one rook on any of the four squares. However, we can rotate the board to obtain any of these from the others. So there is just one distinct configuration. \\
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/one_rook.pdf}}
\]
\item[Two rooks:] We can put rooks on opposite corners. There are two such pairs, but we can rotate the board and obtain either pair from the other. So, again, there is only one distinct possibility.
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/two_rooks.pdf}}
\]
\end{description}
So there are a total of three distinct arrangements of rooks. Now we can try to solve this problem using $\lambda$.

@@

Luckily, we know how to count orbits. All we have to do is find $f$. We go through the subgroups of $C_4$.
\begin{itemize}
\item $C_4$ only fixes the empty board. Any other arrangement can be changed by a rotation. So $f(1) = 1$.
\item The subgroup $\langle \alpha^2 \rangle$ contains the identity element as well as the $180^\circ$ rotation. This subgroup fixes both arrangements with two rooks, as well as the empty board. So $f(2) = 3$.
\item The subgroup $\langle e \rangle$ fixes all the arrangements. There are 2 ways to place two rooks, 4 ways to place one rook, and 1 way to place zero rooks. So $f(2) = 7$.
\end{itemize}
Then all we have to do is compute $\lambda f^T$.
\begin{align*}
\lambda f^T &= \frac 14 [2, 1, 1] \begin{bmatrix} 1 \\ 3 \\ 7\end{bmatrix} \\
&= \frac 14 12\\
&=3.
\end{align*}
So we indeed get the correct answer. 
\end{examp}

In this example, it's not clear that using $\lambda$ actually saved work - in fact, it seemed to take longer. But if we were to increase $n$, we could still determine the number of distinct arrangements relatively easily, because $\lambda$ wouldn't change. Furthermore, in many situations (including this one) it is much easier to determine how many \emph{total} arrangements are fixed by a subgroup that it is to count the number of \emph{distinct} arrangements. We can look at another example, which is more difficult to solve by enumeration. 
\begin{examp}[Example \ref{ex:coloring_init}, cont]\label{ex:coloring}
We now have the tools to solve the problem from Example \ref{ex:coloring_init}. We briefly restate the setup: Suppose we have a $4 \times 4$ board, and we would like to color each square one of $k$ colors. We would like to consider colorings identical if one can be obtained from the other by rotating the board. The goal is to count the number of distinct colorings.

@@

\begin{chapter}{Inclusion-Exclusion}\label{chap:iep}
\section{Introduction}
We start by introducing the desired end result, the \emph{inclusion-exclusion principle}:
\begin{thm}[Inclusion-Exclusion]\label{iep}
Let $S$ be a finite set. Let $A_i \subseteq S$ for each $i \in I$ where $I$ is finite. Let $K \subseteq I$ be given. Then
\[
\left|\bigcup_{i \in K}A_i\right| = \sum_{\emptyset \neq J \subseteq K}(-1)^{|J| + 1}\left|\bigcap_{i \in J}A_j\right|
.\]
\end{thm}
This is a well-known fact. It allows us to find the size of unions of overlapping sets if we have information about intersections of those sets. There are many ways of proving the Inclusion-Exclusion Principle. We will prove it using M\"obius inversion. What is important here is not the result itself -- Theorem \ref{iep} is not hard to prove -- but establishing a connection. In fact, the Inclusion-Exclusion principle is in some sense a special case of M\"obius inversion, and the proof we walk through in this chapter is one way of looking at that connection.

@@

The goal of the Inclusion-Exclusion Principle is, given a subset $K$ of $I$, to find the size of the union of the sets $A_i$ for $i \in K$. The difficulty arises from counting elements multiple times. For example, if we knew that the $A_i$ were disjoint, we could just consider
\[
\sum_{i \in K}\left|A_i\right|.
\]
But, because intersections of the $A_i$ might be nonempty, this sum might count some elements multiple times. We must approach the problem more carefully. First, we define the following indicator function $\chi: s \rightarrow \ps(I)$ by:
\[
\chi(x) = \{i \in I : x \in A_i\}.
\]
So $\chi$ gives us the indices of the subsets of $S$ which contain $x$.  
Next, we define some functions defined on our poset, $\ps(I)$.
Let $J \subseteq I$ be given. Define $f,g,\hat{f}, \hat{g}:\ps(I) \rightarrow \Nn$ by

@@

Also, we define the \emph{complement} function/matrix $c$ as follows: 
\[
c(J,K) = \begin{cases} 1 & J = K^c \\ 0 & \text{otherwise}. \end{cases}
\]
Note that by definition, $c(J,K) = c(K,J)$. So $c$ is symmetric. We also observe that $c^2 = I$, so $c$ is self-inverse.

@@

It makes sense that we would use M\"obius inversion to prove the Inclusion-Exclusion principle. The difficult aspect of counting how many elements were in the union of the sets was avoiding counting elements multiple times. Given a particular set of subsets of $S$, it was easy to determine how many elements were in all of them, because we had access to intersections of the subsets. But what we really needed to know was how many elements were in those subsets and \emph{only} those subsets.

This was a similar problem to what we confronted in Chapter \ref{chap:groups}. There, we were easily able to find elements of the set which were fixed by at \emph{least} some subgroup, but we needed M\"obius inversion to find the precise subgroup that fixed each element, so we could properly count orbits.

In this chapter, we used M\"obius inversion to determine precisely which subsets of $S$ the various set elements were in. This allowed us to determine exactly how many times each element was being counted, and then use facts about binomial coefficients to cancel those extra counts. That was what we did when we found $g$ using $f$ -- the step involving M\"obius inversion. The rest of the proof was simply rearranging terms to arrive at precisely the Inclusion-Exclusion Principle.

@@

\begin{chapter}{Categories and Euler Characteristics}\label{chap:euler}
\section{Introduction to Categories} \label{cat_defs}

@@

We present several examples.
\begin{examp}\label{ex:grp_cat}
Let $G$ be a group. Then we can represent $G$ as a category, which we write as $\star^G$:
\begin{itemize}
\item Let $(\star^G)_0 = \{\star\}$, a single object
\item Let $C(\star, \star) = G$. In other words, let the set of morphisms from $\star$ to $\star$ be equal to the set of group elements.
\item Use the group operation for composition. In other words, if $g,h \in C(\star,\star)$, then, because $g,h \in G$, we can write $gh$. So let $g \circ h = gh$. 
\end{itemize}
It is straightforward to check that $\star^G$ is indeed a category. 
\end{examp}

@@

\begin{defn}\label{def_weighting}
A \emph{weighting} on a category $C$ is a function $\omega$ on $C_0$, the objects of $C$ such that for all $a \in C$,
\[
\sum_{b \in C}|C(a,b)|\omega(b) = 1,
\]
or, equivalently, if
\[
\sum_{b \in C}\zeta(a,b)\omega(b) = 1.
\]
A \emph{coweighting} on $C$ is a function $\kappa$ on $C_0$ such that for all $b \in C$, 
\[
\sum_{a \in C}\kappa(a)|C(a,b)| = 1,
\]
or, equivalently, if
\[
\sum_{a \in C}\kappa(a)\zeta(a, b) = 1.
\]
\end{defn}

One of the first questions we ask when we see this definition is how we can write find valid weightings/coweightings. Luckily, this definition looks familiar. We have a value we know (1) expressed as the sum of function we are seeking ($\omega$). So we can use M\"obius inversion! The situation is slightly more complicated, because we don't know for sure that $\zeta$ is invertible. We formalize that:

\begin{lemma}
\label{mobius_is_weighting}
If $C$ has M\"obius inversion, then the function $\omega_\mu$ defined by $\omega_\mu(a) = \sum_b \mu(a,b)$ is a weighting on $C$. The function $\kappa_\mu$ defined by $\kappa_\mu(b) = \sum_a \mu(a,b)$ is a coweighting on $C$.
\end{lemma}
\begin{proof} For all $a \in C$,
\begin{align*}
\sum_b \zeta(a,b)\omega_\mu(b) &= \sum_b\left( \zeta(a,b)\sum_c \mu(b,c)\right)\\
&=\sum_b\sum_c\zeta(a,b)\mu(b,c)\\
&=\sum_c\left(\sum_b\zeta(a,b)\mu(b,c)\right)\\
&=\sum_c \delta(a,c)\\
&=1.
\end{align*}
The proof is almost identical for $\kappa_\mu$.
\end{proof}

@@

Note that even if $C$ does not admit M\"obius inversion, it might admit a weighting and/or a coweighting. For example, the category in Example \ref{2_pt} admits the weighting and coweighting $\omega(a) =\kappa(a)= \omega(b) = 0.5$, even though it does not admit M\"obius inversion.

\begin{lemma}\label{lem:weight_eq}
Let $C$ be a finite category. Suppose that $\omega$ and $\kappa$ are a weighting and a coweighting respectively on $C$. Then 
\[\sum_{b \in C} \omega(b) = \sum_{a \in C} \kappa(a).
\]
\end{lemma}
\begin{proof}
\begin{align*}
\sum_{b \in C} \omega(b) &= \sum_{b \in C}\left( \sum_{a \in C} \kappa(a)\zeta(a,b)\right)\omega(b) \\
&= \sum_{a \in C}\left(\sum_{b \in C} \zeta(a,b)\omega(b)\right)  \kappa(a) \\
&= \sum_{a \in C}\kappa(a).
\end{align*}
\end{proof}

\begin{defn}[Leinster, \cite{Lein1}]\label{cat_euler}
$C$ be a category with a weighting $\omega$ and a coweighting $\kappa$. Then the \emph{Euler--Leinster characteristic}, or just the \emph{Euler characteristic}, of $C$, denoted by $\chi(C)$, is defined by
\[
\chi(A) = \sum_{b \in C} \omega(b) = \sum_{a \in C} \kappa(a).
\]
\end{defn}
Lemma \ref{lem:weight_eq} tells us that which weighting or coweighting we choose does not change the value of the Euler characteristic.

Together with Lemma \ref{mobius_is_weighting}, we have the following:
\begin{cor}\label{euler_mu}
If a finite category $A$ has a M\"obius inversion $\mu$, then
\[\chi(A) = \sum_{a \subseteq A} \sum_{b \subseteq A} \mu(a, b).
\]
\end{cor}
%TODO: Think of category w/ no euler char?
\section{Simplicial Complexes}
Calling a new construction an Euler characteristic needs some intuitive justification. We will introduce a structure called a \textit{simplicial complex} to show that calling this value an Euler characteristic is reasonable.

@@

The notation for $\cat(\sigma)$ is a little tricky. If $\sigma$ is a simplicial complex, and $\alpha \in \sigma$ is a simplex, then we write $\cat(\alpha)$ to denote the object in $\cat(\sigma)$ corresponding with the simplex $\alpha \in \sigma$. We consider an example.
\begin{examp}
Let $\sigma = \{a, b, ab\}$. Then $\cat(\sigma)$ has three objects and five morphisms, including the three identity morphisms. See Figure \ref{fig:catsigma}.
\begin{figure}[h]
\[
\includegraphics{../Figures/simp_cat.pdf}
\]
\caption{$\sigma$ and $\cat(\sigma)$. Recall that we omit identity morphisms when drawing a category.}
\label{fig:catsigma}
\end{figure}
\end{examp}

@@

There are many methods of hierarchal clustering. In the above example, explicit clusters are formed by grouping points which are near each other. Methods like this are common; see \cite{Carl1}. We will explore an alternate method of clustering, in which we do not explicitly group points, but instead estimates the number of clusters using a real, rather than integer, value. We will use a generalization of the Euler--Leinster characteristic from Chapter \ref{chap:euler}. We start by building some important definitions.

@@

We turn to several examples of monoidal categories.
\begin{examp}
The category $\catname{set}$ is the prototypical example of a monoidal category. In that case, the monoidal product $\otimes$ is the Cartesian product on sets (hence the notation $\otimes$ for the monoidal product), and $u = \{\star\}$, some set with a single element.
